---
title: "DDk_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DDk_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
  - \usepackage[ruled,vlined,linesnumbered]{algorithm2e}
  - \usepackage{fancyhdr,amsmath,amssymb}
  - \usepackage{algorithmic}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(DDK)
```


This is an R implementation for the data driven orthogonal basis selection for functional data analysis paper. For more detaiils see [the arxiv link](https://arxiv.org/pdf/2103.07453.pdf)

In this work, we propose machine learning style techniques for the placement of the knots. The chosen knots are used to build orthogonal splines basis functions $f_k(t), k=1,...,K$ that are used in basis function expansion to convert the data from discrete recorded data into a functional one. The method address the problem of the choice of the initial functional basis selection for functional data analysis. 

\section{DDK algorithm}

# DKK algorithm


\begin{algorithm}[H]
\caption{DDK algorithm}
\label{DDK_Algorithm}
\SetAlgoLined
\KwIn{$I$ - knot search interval, $\theta$ - validation hyperparameter, $\mathcal X_{train}$ - training data, $\mathcal X_{valid}$ - validation data}
 \begin{itemize}
     \item Find the first global knot $\xi$ using equation \ref{new} on $\mathcal X_{train}$\; 
     \item Add split to $I$ at the location of $\xi$ and add $\xi$ to $\mathcal{K}^{(0)}$, the initial set of knots\;
     \item Set $s=1$
 \end{itemize}
 \While {condition \ref{TestTheta_1} or \ref{TestTheta_2} on $\mathcal X_{valid}$ is satisfied}{
 \begin{itemize}
     \item Find the new optimal placement using equation \ref{nominated};
     \item Add split to $I$ at the location of the new selected knot $\xi_s$.
     \item $s = s+1$
 \end{itemize}
 }
\KwOut{$\mathcal{K}^{(s)}$}
\end{algorithm}

## Installation

You can install the development version from [GitHub](https://github.com/ranibasna/ddk) with:

``` r
# install.packages("devtools")
devtools::install_github("ranibasna/ddk")
```




